Biogeography of crop progenitors and wild plant resources in the terminal Pleistocene and Early Holocene of West Asia, 14.7--8.3 ka

### 2025-07-02 analysis/paper.qmd
Update in response to reviews

```{r data-flora}
# Start with flora present at archaeological sites
if (!file_exists(derived_data("flora.Rdata"))) {
  flora <- read_tsv(
    raw_data("swasia_neolithic_flora.tsv"),
    col_types = cols(
      source = col_character(),
      source_id = col_character(),
      source_site_name = col_character(),
      site_name = col_character(),
      latitude = col_double(),
      longitude = col_double(),
      phase = col_character(),
      phase_description = col_character(),
      phase_code = col_character(),
      age_start = col_integer(),
      age_end = col_integer(),
      taxon_source = col_character(),
      n = col_double(),
      prop = col_double(),
      reference = col_character(),
      taxon_detail = col_character(),
      taxon = col_character(),
      genus = col_character(),
      family = col_character(),
      category = col_character(),
      founder_crop = col_character(),
      edibility = col_character(),
      grass_type = col_character(),
      legume_type = col_character()
    )
  ) |>
    rename(taxon_group = "taxon", taxon = "taxon_detail") |>
    # Assign to archaeological periods
    mutate(
      age_mid = age_end + ((age_start - age_end) / 2),
      period = cut(
        -age_mid, 
        breaks = -c(archaeo_periods$start_bp, 0), 
        labels = archaeo_periods$period,
        ordered_result = TRUE
      ),
      climate_period = cut(
        -age_mid, 
        breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
        labels = climate_periods[climate_periods$code != "cur",]$code,
        ordered_result = TRUE
      )
    ) |>
    # Filter to Epipal/Neo
    filter(period %in% epipal | period %in% ppn) |>
    # Group by taxon
    drop_na(taxon) |>
    nest(.by = c(taxon), .key = "archaeo") |>
    # Normalise species names and match to the GBIF Backbone Taxonomy
    mutate(
      # Spot corrections
      taxon = case_match(
        taxon,
        "Vicia narbonense" ~ "Vicia narbonensis", # N=3
        "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
        "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
        "Lens orientalis" ~ "Vicia orientalis", # N=2
        "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
        "Stipa capensis" ~ "Stipa dregeana", # N=2
        .default = taxon
      ),
      taxon_original = taxon,
      taxon = map_chr(taxon, gbif_accepted_name),
      .before = archaeo
    ) |>
    # Drop unmatched (mostly not identified to species)
    drop_na(taxon) |>
    # Recode domesticates as their wild progenitors.
    mutate(
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      )
    ) |>
    # Regroup by canonical taxon
    summarise(
      taxon_aliases = list(taxon_original[taxon_original != taxon]),
      archaeo = list(bind_rows(archaeo)),
      .by = taxon
    ) |>
    mutate(
      n_archaeo = map_int(archaeo, nrow)
    ) |>
    filter(n_archaeo > min_n_archaeo) |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
      )
    )

@@

Taxonomic names were resolved to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023].
So for example occurrences for *Bolboschoenus maritimus* also include those recorded under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

## Occurrence data

@@

@@

# Data preprocessing recipe
# Following recommended preprocessing steps for RF classification from:
#   https://www.tmwr.org/pre-proc-table.html
niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
  update_role(gbif_key:latitude, new_role = "ID") |>
  update_role(present, new_role = "outcome") |>
  update_role(starts_with("bio"), new_role = "predictor") |>
  update_role(c(elevation, slope, aspect, wetness, clay, phh2o, sand, silt),
              new_role = "predictor") |>
  # Impute missing values based on spatial autocorrelation
  step_impute_linear(all_numeric_predictors(), 
                     impute_with = imp_vars(latitude, longitude)) |>
  # Drop redundant predictors
  step_nzv(all_predictors())
```

@@

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on our 24 predictor variables (@sec-predictors).
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (¾) and test (¼) partitions.

@@

Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation (@fig-palaeodist-barely-levant).
It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was managed in the Neolithic.
Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end [@ArranzOtaeguiRoe2023].
We hypothesise that climate-linked range shifts were a factor driving these changes in the apparent economic important of different crops:
the decreasing availability of barley and pistachio in the terminal Pleistocene may have prompted cultivation as a strategy to retain access to them,
though in the long run it made them less attractive as staple crops compared to the more resilient wheat.
However, these dynamics are not seen in the majority of crop progenitors and may also have involved other factors, such as changes in cultural food preferences.

```{r fig-palaeodist-bread}
#| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
bread <- c("Aegilops tauschii", "Aegilops speltoides", 
           "Triticum turgidum dicoccum", "Triticum urartu")
palaeodist_bread <- do.call(c, c(
  filter(flora, taxon %in% bread)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% bread)$taxon))
)) |>
  st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
  mutate(present = factor(present))

@@

Bread wheat (*Triticum aestivum*), the most common wheat cultivar today, has a complex ancestry that involves two recent hybridisation events [@LevyFeldman2022]:
most recently between domestic emmer (*Triticum turgidum dicoccum*) and a goatgrass (*Aegilops tauschii*) c. 9 ka, and before that, in emmer, between wild red einkorn (*T. urartu*) and another goatgrass (*Aegilops speltoides*).
Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to each other in the time frame of domestication is a narrow area around the Orontes river in the northern Levant (@fig-palaeodist-bread).
This area is c. 200 km from the only two sites in our archaeobotanical database where either both species of wheat (*T. urartu* and *T. turgidum dicoccum*) or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is likely to occur in this vicinity.

@@

>>>>

```{r data-flora}
# Start with flora present at archaeological sites
if (!file_exists(derived_data("flora.Rdata"))) {
  flora <- read_tsv(
    raw_data("swasia_neolithic_flora.tsv"),
    col_types = cols(
      source = col_character(),
      source_id = col_character(),
      source_site_name = col_character(),
      site_name = col_character(),
      latitude = col_double(),
      longitude = col_double(),
      phase = col_character(),
      phase_description = col_character(),
      phase_code = col_character(),
      age_start = col_integer(),
      age_end = col_integer(),
      taxon_source = col_character(),
      n = col_double(),
      prop = col_double(),
      reference = col_character(),
      taxon_detail = col_character(),
      taxon = col_character(),
      genus = col_character(),
      family = col_character(),
      category = col_character(),
      founder_crop = col_character(),
      edibility = col_character(),
      grass_type = col_character(),
      legume_type = col_character()
    )
  ) |>
    rename(taxon_group = "taxon", taxon = "taxon_detail") |>
    # Assign to archaeological periods
    mutate(
      age_mid = age_end + ((age_start - age_end) / 2),
      period = cut(
        -age_mid, 
        breaks = -c(archaeo_periods$start_bp, 0), 
        labels = archaeo_periods$period,
        ordered_result = TRUE
      ),
      climate_period = cut(
        -age_mid, 
        breaks = -c(climate_periods[climate_periods$code != "cur",]$start_bp, 0), 
        labels = climate_periods[climate_periods$code != "cur",]$code,
        ordered_result = TRUE
      )
    ) |>
    # Filter to Epipal/Neo
    filter(period %in% epipal | period %in% ppn) |>
    # Group by taxon
    drop_na(taxon) |>
    nest(.by = c(taxon), .key = "archaeo") |>
    # Normalise species names and match to the GBIF Backbone Taxonomy
    mutate(
      # Spot corrections
      taxon = case_match(
        taxon,
        c("Bolboschoenus glaucus", "Bolboschoenus maritimus") ~ "Bolboschoenus",
        "Vicia narbonense" ~ "Vicia narbonensis", # N=3
        "Triticum aestivocompactum" ~ "Triticum aestivum compactum", # N=3
        "Heliotropium persicum" ~ "Moltkia angustifolia", # N=2
        "Lens orientalis" ~ "Vicia orientalis", # N=2
        "Scirpus tabernaemontani" ~ "Schoenoplectus tabernaemontani", # N=2
        "Stipa capensis" ~ "Stipa dregeana", # N=2
        .default = taxon
      ),
      taxon_original = taxon,
      taxon = map_chr(taxon, gbif_accepted_name),
      .before = archaeo
    ) |>
    # Drop unmatched (mostly not identified to species)
    drop_na(taxon) |>
    # Recode domesticates as their wild progenitors.
    mutate(
      taxon = case_match(
        taxon,
        "Cicer arietinum" ~ "Cicer reticulatum",
        "Hordeum vulgare" ~ "Hordeum spontaneum",
        "Vicia lens" ~ "Vicia orientalis",
        "Linum usitatissimum" ~ "Linum bienne",
        "Triticum aestivum" ~ "Triticum turgidum dicoccum",
        "Triticum dicoccoides" ~ "Triticum turgidum dicoccum",
        "Triticum turgidum durum" ~ "Triticum turgidum dicoccum",
        "Triticum monococcum" ~ "Triticum monococcum aegilopoides",
        .default = taxon
      )
    ) |>
    # Regroup by canonical taxon
    summarise(
      taxon_aliases = list(taxon_original[taxon_original != taxon]),
      archaeo = list(bind_rows(archaeo)),
      .by = taxon
    ) |>
    mutate(
      n_archaeo = map_int(archaeo, nrow)
    ) |>
    filter(n_archaeo > min_n_archaeo) |>
    # Add wild progenitors of wheats
    bind_rows(
      tibble(
        taxon = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii"),
        taxon_original = c("Triticum urartu", "Aegilops speltoides", "Aegilops tauschii")
      )
    )

@@

The taxonomic identifications of archaeobotanical material given in our source databases were controlled to ensure consistency between sources and to group taxa that cannot always be reliably distinguished [for details see @ArranzOtaeguiRoe2023].
Taxonomic names were then matched to the canonical form specified in the GBIF Backbone Taxonomy [@GBIFSecretariat2023] so they could be related to modern occurrences.
The aim in both cases was to resolve taxonomic ambiguity by using a more inclusive taxonomic unit.
For example, the model discussed here as *Bolboschoenus* includes both archaeological occurrences of *B. maritimus* and *B. glaucus*, since these were not distinguished by archaeobotanists until recently [@WollstonecroftEtAl2011], as well as modern occurrences under the older nomenclature *Scirpus maritimus* (see @tbl-results-summary).
Domestic species meeting our inclusion criteria were substituted with their wild progenitor(s), where different.

## Occurrence data {#sec-occ-data}

@@

GBIF is currently the best available general-purpose occurrence dataset for the West Asia region, its coverage is uneven both geographically and from species to species.
@fig-occ-map shows clearly that the Southern Levant, and Israel specifically, is significantly more densely sampled than other regions.

@@

# Data preprocessing recipe
# Following recommended preprocessing steps for RF classification from:
#   https://www.tmwr.org/pre-proc-table.html
niche_recipe <- recipe(training(flora$occ_split[[1]])) |>
  update_role(gbif_key:latitude, new_role = "ID") |>
  update_role(present, new_role = "outcome") |>
  update_role(starts_with("bio"), new_role = "predictor") |>
  update_role(c(elevation, slope, aspect, wetness, clay, phh2o, sand, silt),
              new_role = "predictor") |>
  # Impute missing values based on spatial autocorrelation
  step_impute_linear(all_numeric_predictors(), 
                     impute_with = imp_vars(latitude, longitude)) |>
  # Drop correlated predictors
  step_corr(all_predictors(), threshold = 0.7) |>
  # Drop redundant predictors
  step_nzv(all_predictors())
```

@@

For each taxon we trained a classification model to predict occurrence (presence or absence/background) based on up to 24 predictor variables (@sec-predictors).
Highly correlated (Pearson's *r*>0.7) predictors were removed on a taxon-by-taxon basis, to mitigate issues of overfitting due to colinearity [@DormannEtAl2013], as were redundant predictors with zero variance.
We used the Random Forest algorithm implemented in the R package 'ranger' [@WrightZiegler2017] and the 'tidymodels' [@tidymodels] framework for data preprocessing and model selection.
To avoid overfitting, we follow @ValaviEtAl2021 in their recommended hyperparameters and use of down-sampling to balance presence and background samples.
Models for each taxon were fit independently, with redundant zero-variance predictors excluded, and assessed based on balanced training (¾) and test (¼) partitions.

@@

Wild barley (*Hordeum spontaneum*) its relatives show a contraction of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation (@fig-palaeodist-barely-levant).
It also sees a marked decline in the archaeobotanical record from the Early PPNA/Early PPNB (where it was amongst the most common taxa) to the Late PPNB and Late Neolithic [@ArranzOtaeguiRoe2023].
Pistachio (*Pistacia atlantica*) shows similar trends, but it is less certain that this species was managed in the Neolithic.
Conversely, the various wild wheat species native to West Asia show almost no response to Pleistocene/Holocene climate change, even within the Levant, and in the archaeobotanical record wheat displays the opposite trend to barley and pistachio – becoming gradually more abundant through the course of the Neolithic and dominant by its end [@ArranzOtaeguiRoe2023].

```{r fig-palaeodist-bread}
#| fig-cap: Combined predicted palaeodistributions of bread wheat progenitors
bread <- c("Aegilops tauschii", "Triticum aestivum", 
           "Triticum turgidum dicoccum")
palaeodist_bread <- do.call(c, c(
  filter(flora, taxon %in% bread)$palaeodist,
  list(along = list(taxon = filter(flora, taxon %in% bread)$taxon))
)) |>
  st_apply(c("x", "y", "period"), sum, na.rm = FALSE) |>
  mutate(present = factor(present))

@@

Bread wheat (*Triticum aestivum*), the most common wheat cultivar today, has a complex ancestry that involves two recent hybridisation events [@LevyFeldman2022]:
most recently between domestic emmer (*Triticum turgidum dicoccum*) and a goatgrass (*Aegilops tauschii*) c. 9 ka, and before that, in emmer, between wild red einkorn (*T. urartu*) and another goatgrass (*Aegilops speltoides*).
Although there is fairly wide overlap between these species today, according to our modelling the only place all four (or even just three) are predicted to have coincided or been found in close proximity to each other in the time frame of domestication is a narrow area around the Orontes river in the northern Levant (@fig-palaeodist-bread).
This area is c. 200 km from the only two sites in our archaeobotanical database where bread wheat (*T. aestivum*) co-occurs with both of its recent progenitors (*A. tauschii* and *T. turgidum dicoccum*) or both species of wheat and the goatgrasses are found together: Tell Abu Hureyra and El Kowm II [@ArranzOtaeguiRoe2023].^[We are grateful to Benjamin Nowak for pointing this out to us.]
Taken together this suggests an origin of domesticated bread wheat, otherwise only loosely geographically constrained to the Levant–Upper Euphrates corridor [@LevyFeldman2022], is possibly occurred in this vicinity.

@@

For a variety of reasons, our models almost certainly underestimate the *fundamental* niche or potential distribution of the target taxon.
These include:

* The uneven coverage of the GBIF occurrence dataset (see @sec-occ-data)

### 2025-07-02 analysis/paper.tex
Update in response to reviews

Taxonomic names were resolved to the canonical form specified in the
GBIF Backbone Taxonomy \citep{GBIFSecretariat2023}. So for example
occurrences for \emph{Bolboschoenus maritimus} also include those
recorded under the older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where
different.

\subsection{Occurrence data}\label{occurrence-data}

@@

@@

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on our 24 predictor variables
(Section~\ref{sec-predictors}). We used the Random Forest algorithm
implemented in the R package `ranger' \citep{WrightZiegler2017} and the
`tidymodels' \citep{tidymodels} framework for data preprocessing and
model selection. To avoid overfitting, we follow \citet{ValaviEtAl2021}
in their recommended hyperparameters and use of down-sampling to balance
presence and background samples. Models for each taxon were fit
independently, with redundant zero-variance predictors excluded, and
assessed based on balanced training (¾) and test (¼) partitions.

@@

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
decline in the archaeobotanical record from the Early PPNA/Early PPNB
(where it was amongst the most common taxa) to the Late PPNB and Late
Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
atlantica}) shows similar trends, but it is less certain that this
species was managed in the Neolithic. Conversely, the various wild wheat
species native to West Asia show almost no response to
Pleistocene/Holocene climate change, even within the Levant, and in the
archaeobotanical record wheat displays the opposite trend to barley and
pistachio -- becoming gradually more abundant through the course of the
Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}. We
hypothesise that climate-linked range shifts were a factor driving these
changes in the apparent economic important of different crops: the
decreasing availability of barley and pistachio in the terminal
Pleistocene may have prompted cultivation as a strategy to retain access
to them, though in the long run it made them less attractive as staple
crops compared to the more resilient wheat. However, these dynamics are
not seen in the majority of crop progenitors and may also have involved
other factors, such as changes in cultural food preferences.

@@

Bread wheat (\emph{Triticum aestivum}), the most common wheat cultivar
today, has a complex ancestry that involves two recent hybridisation
events \citep{LevyFeldman2022}: most recently between domestic emmer
(\emph{Triticum turgidum dicoccum}) and a goatgrass (\emph{Aegilops
tauschii}) c.~9 ka, and before that, in emmer, between wild red einkorn
(\emph{T. urartu}) and another goatgrass (\emph{Aegilops speltoides}).
Although there is fairly wide overlap between these species today,
according to our modelling the only place all four (or even just three)
are predicted to have coincided or been found in close proximity to each
other in the time frame of domestication is a narrow area around the
Orontes river in the northern Levant
(Figure~\ref{fig-palaeodist-bread}). This area is c.~200 km from the
only two sites in our archaeobotanical database where either both
species of wheat (\emph{T. urartu} and \emph{T. turgidum dicoccum}) or
both species of wheat and the goatgrasses are found together: Tell Abu
Hureyra and El Kowm II \citep{ArranzOtaeguiRoe2023}.\footnote{We are
  grateful to Benjamin Nowak for pointing this out to us.} Taken
together this suggests an origin of domesticated bread wheat, otherwise
only loosely geographically constrained to the Levant--Upper Euphrates
corridor \citep{LevyFeldman2022}, is likely to occur in this vicinity.

@@

>>>>

The taxonomic identifications of archaeobotanical material given in our
source databases were controlled to ensure consistency between sources
and to group taxa that cannot always be reliably distinguished
\citep[for details see][]{ArranzOtaeguiRoe2023}. Taxonomic names were
then matched to the canonical form specified in the GBIF Backbone
Taxonomy \citep{GBIFSecretariat2023} so they could be related to modern
occurrences. The aim in both cases was to resolve taxonomic ambiguity by
using a more inclusive taxonomic unit. For example, the model discussed
here as \emph{Bolboschoenus} includes both archaeological occurrences of
\emph{B. maritimus} and \emph{B. glaucus}, since these were not
distinguished by archaeobotanists until recently
\citep{WollstonecroftEtAl2011}, as well as modern occurrences under the
older nomenclature \emph{Scirpus maritimus} (see
Table~\ref{tbl-results-summary}). Domestic species meeting our inclusion
criteria were substituted with their wild progenitor(s), where
different.

\subsection{Occurrence data}\label{sec-occ-data}

@@

GBIF is currently the best available general-purpose occurrence dataset
for the West Asia region, its coverage is uneven both geographically and
from species to species. Figure~\ref{fig-occ-map} shows clearly that the
Southern Levant, and Israel specifically, is significantly more densely
sampled than other regions.

@@

For each taxon we trained a classification model to predict occurrence
(presence or absence/background) based on up to 24 predictor variables
(Section~\ref{sec-predictors}). Highly correlated (Pearson's
\emph{r}\textgreater0.7) predictors were removed on a taxon-by-taxon
basis, to mitigate issues of overfitting due to colinearity
\citep{DormannEtAl2013}, as were redundant predictors with zero
variance. We used the Random Forest algorithm implemented in the R
package `ranger' \citep{WrightZiegler2017} and the `tidymodels'
\citep{tidymodels} framework for data preprocessing and model selection.
To avoid overfitting, we follow \citet{ValaviEtAl2021} in their
recommended hyperparameters and use of down-sampling to balance presence
and background samples. Models for each taxon were fit independently,
with redundant zero-variance predictors excluded, and assessed based on
balanced training (¾) and test (¼) partitions.

@@

Wild barley (\emph{Hordeum spontaneum}) its relatives show a contraction
of their predicted ranges from the Pleistocene to the Holocene,
concurrent with it being brought into cultivation
(Figure~\ref{fig-palaeodist-barely-levant}). It also sees a marked
decline in the archaeobotanical record from the Early PPNA/Early PPNB
(where it was amongst the most common taxa) to the Late PPNB and Late
Neolithic \citep{ArranzOtaeguiRoe2023}. Pistachio (\emph{Pistacia
atlantica}) shows similar trends, but it is less certain that this
species was managed in the Neolithic. Conversely, the various wild wheat
species native to West Asia show almost no response to
Pleistocene/Holocene climate change, even within the Levant, and in the
archaeobotanical record wheat displays the opposite trend to barley and
pistachio -- becoming gradually more abundant through the course of the
Neolithic and dominant by its end \citep{ArranzOtaeguiRoe2023}.

@@

Bread wheat (\emph{Triticum aestivum}), the most common wheat cultivar
today, has a complex ancestry that involves two recent hybridisation
events \citep{LevyFeldman2022}: most recently between domestic emmer
(\emph{Triticum turgidum dicoccum}) and a goatgrass (\emph{Aegilops
tauschii}) c.~9 ka, and before that, in emmer, between wild red einkorn
(\emph{T. urartu}) and another goatgrass (\emph{Aegilops speltoides}).
Although there is fairly wide overlap between these species today,
according to our modelling the only place all four (or even just three)
are predicted to have coincided or been found in close proximity to each
other in the time frame of domestication is a narrow area around the
Orontes river in the northern Levant
(Figure~\ref{fig-palaeodist-bread}). This area is c.~200 km from the
only two sites in our archaeobotanical database where bread wheat
(\emph{T. aestivum}) co-occurs with both of its recent progenitors
(\emph{A. tauschii} and \emph{T. turgidum dicoccum}) or both species of
wheat and the goatgrasses are found together: Tell Abu Hureyra and El
Kowm II \citep{ArranzOtaeguiRoe2023}.\footnote{We are grateful to
  Benjamin Nowak for pointing this out to us.} Taken together this
suggests an origin of domesticated bread wheat, otherwise only loosely
geographically constrained to the Levant--Upper Euphrates corridor
\citep{LevyFeldman2022}, is possibly occurred in this vicinity.

@@

For a variety of reasons, our models almost certainly underestimate the
\emph{fundamental} niche or potential distribution of the target taxon.
These include:

\begin{itemize}
\tightlist
\item
  The uneven coverage of the GBIF occurrence dataset (see
  Section~\ref{sec-occ-data})
\end{itemize}
